---
title: "Getting Started with mermaid2drawml"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mermaid2drawml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE   # all chunks require Node.js + setup_mermaid2drawml()
)
```

```{r setup}
library(mermaid2drawml)
```

## Introduction

`mermaid2drawml` converts [Mermaid](https://mermaid.js.org/) flowchart diagrams
into natively editable Microsoft Word shapes (DrawingML `wps:wsp` format). Unlike
embedding a static image, the resulting shapes can be selected, moved, resized,
and reformatted directly inside Word.

Layout and SVG rendering are delegated to the official
[Mermaid CLI](https://github.com/mermaid-js/mermaid-cli) (`@mermaid-js/mermaid-cli`),
so Word output matches what you see in a browser. The package integrates with
[officer](https://davidgohel.github.io/officer/), making it easy to insert
diagrams from R Markdown reports.

## Setup

Install Node.js (≥ 18) and npm, then run once after installing the R package:

```{r setup_node}
setup_mermaid2drawml()
```

You can verify the installation at any time:

```{r check_node}
check_mermaid2drawml()
```

## Basic Usage

### `mermaid_to_word()`

The primary function renders Mermaid syntax via the Node.js bridge and builds
DrawingML XML:

```{r basic}
diagram <- mermaid_to_word(
  "flowchart TB
    A[Collect Data] --> B{Quality Check}
    B -->|Pass| C[Analyse]
    B -->|Fail| D[Reject]
    C --> E[Report]"
)

print(diagram)
```

The returned `mermaid_word_diagram` object contains:

- `$xml` — the raw DrawingML XML string ready for `officer::body_add_xml()`
- `$svg_data` — parsed SVG geometry (nodes, edges, viewbox)
- `$ast` — the Mermaid JSON AST from `@mermaid-js/parser`
- `$next_id` — the next available shape ID (needed when adding multiple diagrams)

### `body_add_mermaid()`

For direct insertion into a Word document:

```{r body_add}
library(officer)

doc <- read_docx() |>
  body_add_par("Flowchart of the process", style = "heading 2") |>
  body_add_mermaid(
    "flowchart TB
      A[Start] --> B[Process] --> C[End]"
  )

print(doc, target = "output.docx")
```

## Using in R Markdown

In an R Markdown document, place the insertion code in an R chunk. The output
file must be a Word document (use `output: word_document`):

```
---
output: word_document
---

```{r diagram, echo=FALSE}
library(officer)
library(mermaid2drawml)

my_diagram <- "
flowchart LR
  A[Input] --> B[Transform] --> C[Output]
"

read_docx() |>
  body_add_mermaid(my_diagram) |>
  print(target = knitr::fig_path(".docx"))
```
```

> **Note**: `knitr::fig_path()` generates a unique path per chunk, avoiding
> collisions when multiple diagrams appear in the same document.

## Layout and Sizing Options

`build_diagram_xml()` (called internally) accepts parameters that control how
the SVG is fitted to the page. Pass these through `...` in `mermaid_to_word()`
or `body_add_mermaid()`:

| Parameter | Default | Description |
|-----------|---------|-------------|
| `page_width_in` | 6.0 | Usable diagram width (inches) |
| `page_height_in` | 8.0 | Maximum diagram height (inches) |
| `margin_in` | 1.0 | Page margin offset (inches) |
| `default_fill` | `"FFFFFF"` | Default node fill colour (hex) |
| `default_stroke` | `"4472C4"` | Default stroke colour (hex) |
| `stroke_width_pt` | 1.5 | Stroke width in points |

```{r layout_opts}
diagram <- mermaid_to_word(
  "flowchart LR\n  A --> B --> C",
  page_width_in = 5.0,
  margin_in     = 0.5
)
```

## Multi-Diagram Documents

When adding multiple diagrams to a single document, chain `start_id` / `next_id`
to avoid shape ID collisions (Word requires unique IDs):

```{r multi_diagram}
library(officer)

d1 <- mermaid_to_word(
  "flowchart TB\n  A --> B",
  start_id = 1L
)

d2 <- mermaid_to_word(
  "flowchart LR\n  C --> D --> E",
  start_id = d1$next_id   # pick up where d1 left off
)

doc <- read_docx() |>
  officer::body_add_xml(d1$xml) |>
  body_add_par("") |>
  officer::body_add_xml(d2$xml)

print(doc, target = "multi.docx")
```

## Supported Mermaid Syntax

The package supports the full feature set of `@mermaid-js/mermaid-cli`, including:

- **Directions**: `TB`, `TD`, `LR`, `BT`, `RL`
- **Node shapes**: `[rect]`, `(roundRect)`, `{diamond}`, `((ellipse))`,
  `{{hexagon}}`, `[(cylinder)]`, `@{shape: ...}` (v10 shapes)
- **Edge types**: `-->` (arrow), `---` (line), `-.->` (dashed), `<-->` (bidirectional),
  `o--o` (circle ends), `x--x` (cross ends), `==>` (thick)
- **Edge labels**: `-->|label|` and inline `-- label -->`
- **Chained edges**: `A --> B --> C`
- **Subgraphs**
- **`classDef`** style definitions with `fill`, `stroke`, `color`
- **Comments**: `%% ...`

## Round-Trip Metadata

Each shape and connector in the generated Word document stores JSON metadata in
its `wp:docPr descr` attribute. This metadata records the original Mermaid
properties (node ID, label, shape type, edge endpoints, etc.) under a versioned
schema (`"v": "1"`). A future Word-to-Mermaid conversion feature can use this
metadata to reconstruct the diagram source from an edited Word document.
